---
title: "practical_exercise_1, Methods 3, 2021, autumn semester"
author: 'Daniel'
date: "20-09-21"
output: html_document
---

<style type="text/css">
  body{
  font-size: 14pt;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# Exercises and objectives
The objectives of today's exercises are:  
1) To remind you of the (general) linear model, and how we can use it to make models in R  
2) To make some informal model comparisons  
3) To estimate models based on binomially distributed data  

If you would like to read more about a given function, just prepend the function with a question mark, e.g.  
``` {r, eval=FALSE}
?lm
```

REMEMBER: In your report, make sure to include code that can reproduce the answers requested in the exercises below   

## Exercise 1
The general linear model: $Y = X \beta + \epsilon$:  
Do a linear regression, expressing fuel usage as a function of weight using the function __lm__  
```{r, eval=FALSE}
data(mtcars)
model <- lm(formula=mpg~wt, data=mtcars)
```
1. extract $\hat{\beta}$, $Y$, $\hat{Y}$, $X$ and $\epsilon$ from __model__ (hint: have a look at the function __model.matrix__)  
    i. create a plot that illustrates $Y$ and $\hat{Y}$ (if you are feeling ambitious, also include $\epsilon$ (hint: you can use the function __arrows__))
```{r}
b_hat <- coef(model)
y <- mtcars$mpg
y_hat <- predict(model)
X <- model.matrix(model)
x <- mtcars$wt
epsilon <- model$residuals

# Make a data frame that is ggplot-friendly
df_x <- c(x,x)
df_y <- c(y, y_hat)
df_index <- c(rep(1,32), rep(2,32)) # a column that will indicate whether the y-value is just y or y_hat
df <- as.data.frame(cbind(df_x, df_y, df_index))

ggplot(df, aes(df_x, df_y, color=df_index))+geom_point()+theme_bw()

```

2. estimate $\beta$ for a quadratic model ($y = {\beta}_{2} x^2 + {\beta}_{1} x + {\beta}_{0}$) using ordinary least squares _without_ using __lm__; $\hat{\beta} = {({X}^{T} X)}^{-1} {X}^{T} Y$ (hint: add a third column to $X$ from step 1)
```{r}
X <- cbind(X,x^2)
OLS <- solve(t(X)%*%X)%*%t(X)%*%y
```


3. compare your acquired $\hat{\beta}$ with the output of the corresponding quadratic model created using __lm__ (hint: use the function __I__, see details under help and the sub-section formula operators here: https://www.datacamp.com/community/tutorials/r-formula-tutorial)
    i. create a plot that illustrates $Y$ and $\hat{Y}$ (if you are feeling ambitious, also include $\epsilon$ (hint: you can use the function __arrows__))
```{r}
model2 <- lm(mpg~I(wt^2)+wt, data=mtcars)

# Comparison of model estimates... estimates are identical
OLS
coef(model2)

# creating a plot using a different method than in exercise 1.1
y_predict_squared <- predict(model2)
ggplot(mtcars, aes(wt, mpg))+
  geom_point()+
  geom_point(aes(x,y_predict_squared))+
  stat_smooth(aes(y=y_predict_squared),method = "lm", formula = y ~ x + I(x^2), size = 1, color = "red")
```


## Exercise 2
Compare the plotted quadratic fit to the linear fit  

1. which seems better?
It seems like the quadratic model fits the mpg-values better.

2. calculate the sum of squared errors, (show the calculation based on $\epsilon$). Which fit has the lower sum? 
```{r}
#SSE linear
sum(epsilon^2)

#SSE quadratic, lower than the other
sum(model2$residuals^2)
```

3. now make a cubic fit ($y = {\beta}_{3} x^3 + {\beta}_{2} x^2 + {\beta}_{1} x + {\beta}_{0}$) and compare it to the quadratic fit
```{r}
model3 <- lm(mpg~I(wt^3)+I(wt^2)+wt, data = mtcars)
summary(model3)
```

    i. create a plot that illustrates $Y$ and $\hat{Y}$ for both the cubic and the quadratic fits (plot them in the same plot) 
```{r}
y_predict_cubic <- predict(model3)

ggplot(mtcars, aes(wt, mpg))+
  geom_point()+
  geom_point(aes(x,y_predict_squared))+
  stat_smooth(aes(y = y_predict_squared),method = "lm", formula = y ~ x + I(x^2), size = 1, color = "red")+
  geom_point(aes(x, y_predict_cubic))+
  stat_smooth(aes(y = y_predict_cubic),method = "lm", formula = y ~ x + I(x^2) + I(x^3), size = 1, color = "blue")

```

    ii. compare the sum of squared errors
```{r}
sum(model3$residuals^2) # the SSE is slightly smaller than for the quadratic model. Nevertheless, they are very similar which is consistent with the fact that the graphs are almost completely overlapping.
```

    iii. what's the estimated value of the "cubic" (${\beta}_3$) parameter? Comment on this!
```{r}
coef(model3)[2] # It is quite a small number. So, the cubic term does not have a large influence on our model and therefore we see in ii. that this model does not explain much more variance than the quadratic model.
```
    
4. bonus question: which summary statistic is the fitted value (_Intercept_ or ${\beta}_0$ in $y = {\beta}_0$) below identical to?
```{r, echo=FALSE}
# The model y ~ 1 is equal to computing the mean of y
lm(mpg ~ 1, data=mtcars)
mean(mtcars$mpg)
```


## Exercise 3
Doing a logistic regression - estimating the probability that a car has automatic transmission (0) or manual transmission (1) based on its weight
```{r, eval=FALSE}
data(mtcars)
logistic.model <- glm(formula=am~wt, data=mtcars, family='binomial')
```

Probabilities live on the range $(0, 1)$ - using the so-called logit function as a "link-function" we can map these onto the range $(-\infty, \infty)$, i.e. the real numbers.  
  
What we model in this case is: $Pr(y = 1) = logit^{-1}(X \beta)$, i.e. the probability of a car having manual transmission, given its weight. $X \beta$ is called the linear predictor; compare with $Y = X \beta + \epsilon$ 
It is helpful to define the logit function and its inverse function for the following:  

```{r}
logit <-     function(x) log(x / (1 - x))
inv.logit <- function(x) exp(x) / (1 + exp(x))
```

1. plot the fitted values for __logistic.model__:  
    i. what is the relation between the __linear.predictors__ and the __fitted_values__ of the __logistic.model__ object?
```{r}
# I do not understand this question
```

2. plot the logistic function, you've estimated based on your $\hat{\beta}$, (not just the fitted values). Use an _xlim_ of (0, 7)
```{r}
y_on_prob_scale <- inv.logit(predict(logistic.model))
ggplot(mtcars, aes(wt, y_on_prob_scale))+geom_point()+xlim(0,7)+theme_bw()
```

    i. what's the interpretation of the estimated $\hat{\beta}_0$ (the _Intercept_)
```{r}
# For a hypothetical car, if wt=0, the probability of that car having manual transmission is 
inv.logit(12.040)
```

    ii. calculate the estimated probability that the Pontiac Firebird has automatic transmission, given its weight
```{r}
prob_manual <- inv.logit(12.040-4.024*3.845) # this is the probability that the car has manual transmission. Thus the probability of the car having automatic transmission must be
1-prob_manual
```
    
    iii. bonus question - plot the logistic function and highlight all the cars where we guessed wrongly, if we used the following "quantizer" function:
    
\begin{equation}
  transmission_{guess}=
  \begin{cases}
    1 (manual), & \text{if}\ PR(y = 1) ≥ 0.5 \\
    0 (automatic), & \text{otherwise}
  \end{cases}
\end{equation}    

```{r}
# I make a column in mtcars that indicate whether the predicted transmission is different from the true transmission (note: our quantizer function is equal to rounding our predictions)
mtcars$wrong_guess <- ifelse(mtcars$am != round(y_on_prob_scale), 1, 0)

ggplot(mtcars, aes(wt, y_on_prob_scale, color=wrong_guess))+geom_point()+xlim(0,7)+theme_bw()

```


3. plot quadratic fit alongside linear fit  
    i. judging visually, does adding a quadratic term make a difference?
```{r}
ggplot(mtcars, aes(wt, mpg))+
  geom_point()+
  geom_point(aes(x,y_predict_squared))+
  stat_smooth(aes(y = y_predict_squared),method = "lm", formula = y ~ x + I(x^2), size = 1, color = 'red')+
  geom_point(aes(x, y_hat))+
  stat_smooth(aes(y = y_hat),method = "lm", formula = y ~ x, size = 1, color = 'green')

# Adding a quadratic term seems to make a positive difference, see exercise 2.1
```

    ii. check the details in the help of the AIC function - which of the models provide the better fit according to the AIC values and the residual deviance respectively?
```{r}
AIC(model, model2, model3)

# The quadratic model has the lowest AIC-value, and therefore provides the better fit according to this model evaluation.
```

    iii. in your own words, why might it be good to penalise a model like the quadratic model, we just fitted.
```{r}
# I'm not sure :)
```

